Adaption of the Chumbley Score to matching of bullet striation marks

### 2018-01-03 short/short.Rmd
short version added

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge ...
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- 3 to 6 keywords
- that do not appear in the title
bibliography: bibliography
abstract: null
---

@@

\tableofcontents

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "",
  fig.height = 6,
  fig.width = 6,
  fig.align = "center",
  out.width= '\\textwidth',
  cache = FALSE,
  fig.path='figures/',
  echo=FALSE,
  cache=TRUE
)
options(knitr.table.format = "latex")
library(tidyverse)
```

```{r functions}
errorrate <- function(data, alpha) {
  summ <- data %>% filter(!is.na(p_value)) %>%
    mutate(signif = p_value < alpha) %>%
    group_by(wv, wo, match, coarse, signif) %>% tally()
  summ$error <- with(summ, match != signif)
  summ$alpha <- alpha

  rates <- summ %>% group_by(wv, wo, coarse, match, alpha) %>% summarize(
    rate = n[error==TRUE]/sum(n)
  )
  totals <- summ %>% group_by(wv, wo, coarse, alpha) %>% summarize(
    total= sum(n[error==TRUE])/sum(n)
  )
  rates <- rates %>% spread(match,rate) %>% rename(
    actual = `FALSE`,
    beta = `TRUE`
  )
  left_join(rates, totals)
}

# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
# Multiple plot function
#
# ggplot objects can be passed in ..., or to plotlist (as a list of ggplot objects)
# - cols:   Number of columns in layout
# - layout: A matrix specifying the layout. If present, 'cols' is ignored.
#
# If the layout is something like matrix(c(1,2,3,3), nrow=2, byrow=TRUE),
# then plot 1 will go in the upper left, 2 will go in the upper right, and
# 3 will go all the way across the bottom.
#
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
#  ref: http://www.cookbook-r.com/Graphs/Multiple_graphs_on_one_page_(ggplot2)/
```

```{r data}
if (!file.exists("../data/all-sigs.rds")) {
  files <- dir("../data/signatures", pattern="csv")
  all <- data.frame()
  for (file in files) {
    tmp <- read.csv(file=file.path(path="../data/signatures", file)) 
    if (length(grep("x",names(tmp)) > 0)) tmp <- tmp %>% select(-x)
    all <- rbind(all, tmp)
  }
  all <- all %>% filter(land1_id < land2_id)
  saveRDS(all, file="../data/all-sigs.rds")
} else {
  all <- readRDS("../data/all-sigs.rds")
}

all$coarse <- 1
errors <- rbind(errorrate(all, 0.001), 
                errorrate(all, 0.005), 
                errorrate(all, 0.01), 
                errorrate(all, 0.05))
```

```{r data-profiles}
if (!file.exists("../data/all-profiles.rds")) {
  files <- dir("../data/profiles/", pattern="csv")
  allp <- data.frame()
  for (file in files) {
    tmp <- read.csv(file=file.path(path="../data/profiles", file)) 
    #print(names(tmp))
    if (length(grep("coarse",names(tmp))) == 0)
      tmp$coarse <- gsub(".*coar-(.*).csv", "\\1", file)
    if (length(grep("x",names(tmp)) > 0)) tmp <- tmp %>% select(-x)
    allp <- rbind(allp, tmp)
  }
  allp <- allp %>% filter(land1_id < land2_id)
  allp <- allp %>% mutate(coarse = as.numeric(gsub("pt", ".", coarse)))
  saveRDS(allp, file="../data/all-profiles.rds")
} else {
  allp <- readRDS("../data/all-profiles.rds")
}

errors2 <- rbind(errorrate(allp, 0.001), 
                errorrate(allp, 0.005), 
                errorrate(allp, 0.01), 
                errorrate(allp, 0.05))

```

```{r data-sig-coars}
if (!file.exists("../data/all-sig-c.rds")) {
  files <- dir("../data/sig_diff_coarseness/", pattern="csv")
  all_sig_c <- data.frame()
  for (file in files) {
    tmp <- read.csv(file=file.path(path="../data/sig_diff_coarseness/", file)) 
    #print(names(tmp))
    if (length(grep("coarse",names(tmp))) == 0)
      tmp$coarse <- gsub(".*coarse-(.*).csv", "\\1", file)
    if (length(grep("x",names(tmp)) > 0)) tmp <- tmp %>% select(-x)
    all_sig_c <- rbind(all_sig_c, tmp)
  }
  all_sig_c <- all_sig_c %>% filter(land1_id < land2_id)
  all_sig_c <- all_sig_c %>% mutate(coarse = as.numeric(gsub("pt", ".", coarse)))
  saveRDS(all_sig_c, file="../data/all-sig-c.rds")
} else {
  all_sig_c <- readRDS("../data/all-sig-c.rds")
}

errors_sig_c <- rbind(errorrate(all_sig_c, 0.001), 
                errorrate(all_sig_c, 0.005), 
                errorrate(all_sig_c, 0.01), 
                errorrate(all_sig_c, 0.05))

```

# Introduction and Background

## Problem statement and motivation

Striation marks are used to determine same source for bullets. Current standards ask for a visual inspection of the marks under a comparison microscope by a firearms examiner. One of the problems raised by the NAS report in 2009 is the subjectivity involved in this comparison and the need for determining error rates \citep{NAS:2009}. The issue of subjectivity of a firearms examiner has been reviewed by many authors where things like scientific principle testability and error rates are considered to be defining aspects for an objective analysis such that without identifying the uncertainities associated with a method of comparison, it is inconclusive to regards an analysis complete. This means that with the usual method of visual comparison that is adopted by firearms examiners, lies the problem of lack of any systemized quantification and as such there seems no certain way of associating it with any kind of mathematical probability. Since determining error rates has been duly noted as a fundamental problem in forensic science \citep{NAS:2009}, there is a need of methods that address this problem. Some of the methods that have been used to estimate error rates in firearm examinations include Automatic cartridge case comparison where same source identification of cartridge cases and associated error rates have been provided by \citep{riva}, although as noted by \citep{aoas}, in this case alignments of striae involves roation of planes, which cannot be generalized for bullets. In case of methods that use machine learning algorithms some methods which are bootsrap based methods often tend to give over-estimated error rates \citep{efron}, but there are alternate error estimation techniques as described by \citep{aoas}, \citep{efron} and  \citep{vorburger2016} which give better results.

In case of methods that do not use machine learning algorithm, matching two striation marks with each other in order to identify if it is from the same source, can also be done using a well defined comparative statistical algorithm. The statistical algorithm is therefore expected  to go through a step by step procedure of identifying the class characteristics i.e striation marks for bullets, choosing the striae, and then follow a systematic procedure of comparing two striae with each other on the basis of criteria like cross-correlation (or others). Such a methodology would make it possible to determine error rates too and give definite results regarding what proportion of cases would the analysis be successful in identifying a match correctly and in how many cases would it end up being a false positive.

Identification criteria like consecutively matching striae (CMS) as first seen by Biasotti 1959 \citet{biasotti} and later mentioned in \citet{chu2013} more often than not include subject bias and are error prone and as seen in the work of \citet{miller} the number of CMS may turn out to be high even when the bullet is not fired by the same firearm. Parameters like cross-correlation factor as seen in the extensive comparisons made by \citep{aoas} seem to perform much better for matching purposes.

An important aspect of same source matching in firearms is identification of bullet microtopographies that are unique to be chosen for a match, Bullet profiles and signatures prove to be such features. The choice of a bullet signature to uniquely define a bullet depends on finding first, a stable region on a Bullet land which minimal noise and many pronounced striation marks. \citet{aoas}

The method employed by \citet{aoas} uses the cross correlation factor as a means to identify the stable region. The markings or striae in this region are supposed to have a high CCF with each other. Therefore a Bullet profile is chosen from this region by taking the cross section at a given height. A loess fit on the bullet profile produces  residuals which are termed as signatures and can be used as a unique identifier that can be used while matching two bullets to a firearm.

Compairing pairs of toolmarks on the otherhand, with the intention of matching it to a tool has been studied relatively more in the past as compared to bullets, and \citet{chumbley} have described in their paper an algorithm and analytic method that compares two toolmarks and come to the conclusion if they are from the same tool or not. The method also determines the error rates, reduces subject bias and designate the two toolmarks as matches or non-matches with respect to a source. \citet{chumbley} used an empirical based setup to validate their proposed analytic and quantitative algorithm.

\citep{chumbley} found that the algorithm gives false-positives in slightly over 2 cases (or 2%) of the time and about 9 false-negatives (or 9%) for every 100 comparisons. \citep{hadler} on the other hand, in the improved version of the algorithm first described by \citep{chumbley} found false-positives for 0 cases and 3 false-negatives (or 6%) for 50 comparisons of knwon matches.
\begin{figure}
\centering
\includegraphics[width=\textwidth]{images/B6-B2-L6.png}
```{r, echo=FALSE, fig.width=8, fig.height=3, out.width='\\textwidth'}
land <- read.csv("../data/b6b2l6.csv")
land %>% ggplot(aes(x = y, y=resid)) +geom_line() +theme_bw()
```

\caption{\label{fig:rgl} Image of a bullet land from a confocal light microscope at 20 fold magnification.}

\end{figure}

# The Chumbley Score Test

The Chumbley score algorithm takes input as two vectorized processes $x(t_1)$, $t_1 = 1,2,...T_1$ and $y(t_2)$, $t_2 = 1,2...T_2$ which denote two sets of marks or striae. Here the processes are of the form $z(s)$ which is a spatial process for some $t$, while $z(s_1)$ is the realization of the process in $s_1$. The $t_1$ and $t_2$ can be better understood as equally spaced pixel locations for the two marks under consideration, where a 'pixel' is the resolution of the confocal light microscope. In the case for bullet signatures a pixel corresponds to 0.645 microns. These marks or striae are potentially from two different bullets or two different toolmarks whose source needs to be identified as being same or different. The marks or striae are indexed by the pixel location where $t_1$ is for the first striae referred to as $x$ and $t_2$ is for the second striae which is referred as $y$. $T_1$ and $T_2$ being the respective lengths of the markings, need not be the same but are usually of similar lenghtsThe similarity is then judged by the algorithm on the basis of cross-correlation of a fixed and constant number consecutive pixels say $k$ taken from the two indexed marks $x(t_1)$ and $y(t_2)$ such that in theory $k$ remains smaller that length of the two striae or marks which is the same as $t_1$ and $t_2$. Depending on the what stage of the algorithm we are in, matching of different pixel lengths and locations is done, which at the end effectively compares all possible windows that would guarantee in quantifying the two marks or striae as coming from the same source or not, which also lets us assess the error rates by checking for a large number of cases.

The algorithm works in two phases, namely, an optimization step and a validation step, at the end of which a Mann Whitney U statistic is calculated. A pre-processing step to the algorithm is to choose a coarseness value which  is used as a parameter to the LOWESS smoothing function. The coarseness essentially gives the proportion of points which influence the smooth at each value, which means larger values lead to more smoothness. The LOWESS smoothing is applied to each of two sets of vectorized striae or marks $x(t_1)$ and $y(t_2)$, before proceeding to the algorithm.

\citet{hadler} in their paper proposed an improvement to this algorithm by trying to remove mutual dependence of parameters (due to serial correlation in surface depth values of a toolmark and because of a random sampling sub step in the validation phase which makes a group of pixels to be chosen more than once and hence introduces lack of independence) in certain steps, especially because the Mann-Whitney U statistic that is later calculated in the algorithm and used as a measure to differentiate between matches and non-matches works under the assumption of independence of parameters.

Since we are only interested in a non-parameteric U statistic, Hadler et al. proposed a normalization procedure in the Validation step that goes to some extent to address the issue of mutual dependence. Also the same shift and different shift substep was modified to use a deterministic rule for sampling sam-shhift and dofferent shift-samples as opposed to the originally proposed random samples.

The data that is used by \citet{chumbley} is generated by a surface profilometer that gives the height in terms of distance along a linear trace. This is taken perpendicular to the striations present in the toolmark. Two such trace are then compared using the algorithm.

## Detailed algorithm

###Optimization step
The idea behind this step is to first identify the area of best agreement in the two toolmark data. Comparison window size is predefined by the user, which in case of screwdriver toolmarks was chosen by Chumbley et al and Hadler et al as around 10 percent of the length of the toolmarks, which was around 500. This window is henceforth referred to as Window of optimization.

A maximum correlation statistic is used to identify the region of best agreement, with the maximum usually seen being near 1 for both cases which is what we intuitively expect for matches, and something which we do not intuitively expect for non matches

First, there are a very large number of cross-correlations calculated for the two series of striae or marks $x(t_1)$ and $y(t_2)$, for eg if the window of optimization was defined as 200 and toolmark pixel length is around 1000 then we have 1200-(200-1) = 1001.

This is the number of windows we have for one toolmark and each window is compared with all windows of the second toolmark (the number of windows is again similar), and the window with the maximum correlation is identified to be the region where the toolmarks are in maximum agreement with each other

###Validation Step

####Same-shift: 
In this step a series of windows are chosen at random (originally by chumbley et al) and deterministically (by hadler et al), but at a common distance (rigid-shift) from the window identified as the region of best agreement in the vaildation step.
The correlation of these windows would be as intuitively assumed i.e. lower than the maximum correlation window (Optimization step), but the significance is that these same shift windows will still have large enough correlation values for two toolmarks or signatures that are in reality a match.

This also validates that if in the optimization step the maximum correlation window chosen (with correlation value near 1) was by accident (like in case of signatures that in reality are a non-match), all these same- shift correlations (A fixed number of these trace segments are identified) would not be anywhere large enough for all same-shift windows.

```{r win-comparison, fig.width=8, fig.height=5, out.width='\\textwidth', fig.cap=" The two plots on the left show how the same shift behaves in case of a matching pair and the two plots on the right show how the different shift behaves in case of a matching pair."}
a<- matrix(NA, ncol = 3, nrow = 2)
toolmark.falsepositive<- data.frame(a)
colnames(toolmark.falsepositive)<- c("Classification", "Match", "Non-Match")
toolmark.falsepositive$Classification<- c("Match", "Non-Match")
toolmark.falsepositive$Match<- c(47,0)
toolmark.falsepositive$`Non-Match`<- c(3,50)

knitr::kable(toolmark.falsepositive, digits=0, format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("hold"))

asd<- all %>% filter(wo==120, wv==50, match == TRUE, land1_id == 1)

sigs_graphics<- readRDS("../data/sigs_generate_window.rds")

sigs_graphics.match<- as.data.frame(sigs_graphics[1])
sigs_graphics.nonmatch<- as.data.frame(sigs_graphics[2])
colnames(sigs_graphics.match)<- gsub("^.*\\.","", colnames(sigs_graphics.match))
colnames(sigs_graphics.nonmatch)<- gsub("^.*\\.","", colnames(sigs_graphics.nonmatch))

# Matching signature
#gi<-
#sigs_graphics.match<- sigs_graphics.match %>% filter(land_id == c(1,9))  
#sigs_graphics.match<- sigs_graphics.match %>% select(y, l30, land_id)
d1<- sigs_graphics.match[which(sigs_graphics.match$land_id==1,arr.ind = TRUE),]
d2<- sigs_graphics.match[which(sigs_graphics.match$land_id==19,arr.ind = TRUE),]
data1<-d1 %>%select(l30)
#data1<- data1$y
data2<- d2 %>%select(l30)
#data2<- data2$y
window_opt = 120 
window_val = 50
coarse = 1
data1<- matrix(unlist(data1))
data2<- matrix(unlist(data2))

  unity <- function(x) {x / sqrt(sum(x^2))} ## normalize columns of a matrix to make correlation computation faster

  ####################################################
  ##Clean the marks and compute the smooth residuals##
  ####################################################

  data1 <- matrix(data1[round((0.01*nrow(data1))):round(0.99*nrow(data1)),], ncol = 1)
  data2 <- matrix(data2[round((0.01*nrow(data2))):round(0.99*nrow(data2)),], ncol = 1)

  ##Normalize the tool marks
  y1 <- data1 - lowess(y = data1,  x = 1:nrow(data1), f= coarse)$y
  y2 <- data2 - lowess(y = data2,  x = 1:nrow(data2), f= coarse)$y

  ############################################
  ##Compute the observed maximum correlation##
  ############################################

  #####################
  ##Optimization step##
  #####################
  ##Each column in these matrices corresponds to a window in the respective tool mark
  y1_mat_opt <- matrix(NA, ncol = length(1:(length(y1) - (window_opt - 1))), nrow = window_opt)
  for(l in 1:(length(y1) - (window_opt - 1))){
    y1_mat_opt[,l] <- y1[l:(l+(window_opt - 1))]
  }
  y2_mat_opt <- matrix(NA, ncol = length(1:(length(y2) - (window_opt - 1))), nrow = window_opt)
  for(l in 1:(length(y2) - (window_opt - 1))){
    y2_mat_opt[,l] <- y2[l:(l+(window_opt - 1))]
  }

  ##Compute the correlation between all pairs of windows for the two marks
  ##Rows in the following matrix are mark 2, columns are mark 1
  y2_mat_opt <- apply(scale(y2_mat_opt), 2, unity)
  y1_mat_opt <- apply(scale(y1_mat_opt), 2, unity)
  corr_mat_opt <- t(y2_mat_opt) %*% y1_mat_opt ##correlation matrix
  max_corr_opt_loc <- which(corr_mat_opt == max(corr_mat_opt), arr.ind = TRUE) ##pair of windows maximizing the correlation

  s1.rect.opt <- data.frame(xmin=max_corr_opt_loc[1,2], xmax= max_corr_opt_loc[1,2]+window_opt, ymin=-Inf, ymax=Inf)
    s1.rect.sameshift1 <- data.frame(xmin=max_corr_opt_loc[1,2]-window_val, xmax= max_corr_opt_loc[1,2]- 2*window_val, ymin=-Inf, ymax=Inf)
       s1.rect.sameshift2 <- data.frame(xmin=max_corr_opt_loc[1,2]+window_opt+ window_val, xmax= max_corr_opt_loc[1,2]+ window_opt+2*window_val, ymin=-Inf, ymax=Inf)

  s2.rect.opt <- data.frame(xmin=max_corr_opt_loc[1,1], xmax= max_corr_opt_loc[1,1]-window_opt, ymin=-Inf, ymax=Inf)
    s2.rect.sameshift1 <- data.frame(xmin=max_corr_opt_loc[1,1]-window_opt-window_val, xmax= max_corr_opt_loc[1,1]-window_opt- 2*window_val, ymin=-Inf, ymax=Inf)
       s2.rect.sameshift2 <- data.frame(xmin=max_corr_opt_loc[1,1]+ window_val, xmax= max_corr_opt_loc[1,1]+ 2*window_val, ymin=-Inf, ymax=Inf)

p1<-  ggplot(data = d1, aes(x = y, y = l30)) + geom_line() + 
  geom_rect(data=s1.rect.opt, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "black",alpha=0.5, inherit.aes = FALSE)+
    geom_rect(data=s1.rect.sameshift1, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "red",alpha=0.1, inherit.aes = FALSE)+
    geom_rect(data=s1.rect.sameshift2, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "blue",alpha=0.1, inherit.aes = FALSE)+
  theme_bw() + labs(y = "Signature Land Id = 1")

 p2<-   ggplot(data = d2, aes(x = y, y = l30)) + geom_line() + 
   geom_rect(data=s2.rect.opt, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "black",alpha=0.5, inherit.aes = FALSE)+
    geom_rect(data=s2.rect.sameshift1, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "red",alpha=0.1, inherit.aes = FALSE)+
    geom_rect(data=s2.rect.sameshift2, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "blue",alpha=0.1, inherit.aes = FALSE)+
  theme_bw() + labs(y = "Signature Land Id = 9")
 ##########

 d1.rect.opt <- data.frame(xmin=max_corr_opt_loc[1,2], xmax= max_corr_opt_loc[1,2]+window_opt, ymin=-Inf, ymax=Inf)
    d1.rect.sameshift1 <- data.frame(xmin=max_corr_opt_loc[1,2]-2*window_opt, xmax= max_corr_opt_loc[1,2]-2*window_opt-window_val, ymin=-Inf, ymax=Inf)
       d1.rect.sameshift2 <- data.frame(xmin=max_corr_opt_loc[1,2]+window_opt+ window_val, xmax= max_corr_opt_loc[1,2]+ window_opt+2*window_val, ymin=-Inf, ymax=Inf)

  d2.rect.opt <- data.frame(xmin=max_corr_opt_loc[1,1], xmax= max_corr_opt_loc[1,1]-window_opt, ymin=-Inf, ymax=Inf)
    d2.rect.sameshift1 <- data.frame(xmin=max_corr_opt_loc[1,1]-window_opt-window_val, xmax= max_corr_opt_loc[1,1]-window_opt- 2*window_val, ymin=-Inf, ymax=Inf)
       d2.rect.sameshift2 <- data.frame(xmin=max_corr_opt_loc[1,1]+ 4*window_val, xmax= max_corr_opt_loc[1,1]+ 5*window_val, ymin=-Inf, ymax=Inf)

p3<-  ggplot(data = d1, aes(x = y, y = l30)) + geom_line() + 
  geom_rect(data=d1.rect.opt, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "black",alpha=0.5, inherit.aes = FALSE)+
    geom_rect(data=d1.rect.sameshift1, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "red",alpha=0.1, inherit.aes = FALSE)+
    geom_rect(data=d1.rect.sameshift2, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "blue",alpha=0.1, inherit.aes = FALSE)+
  theme_bw() + labs(y = "Signature Land Id = 1")

 p4<-   ggplot(data = d2, aes(x = y, y = l30)) + geom_line() + 
   geom_rect(data=d2.rect.opt, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "black",alpha=0.5, inherit.aes = FALSE)+
    geom_rect(data=d2.rect.sameshift1, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "red",alpha=0.1, inherit.aes = FALSE)+
    geom_rect(data=d2.rect.sameshift2, aes(xmin=xmin, xmax=xmax , ymin=ymin, ymax=ymax),color = "blue",alpha=0.1, inherit.aes = FALSE)+
  theme_bw() + labs(y = "Signature Land Id = 9")

multiplot(p1, p2, p3, p4, cols=2)
```

####Different Shift:
Primary reason for this substep is to give perspective to the correlation values of the same shift window correlation values.

This time there are no rigid-shifts but different shifts (distance from Window of Opt with max correlation) chosen randomly by \citet{chumbley} and deterministically by Hadler et al. such that there is an equal possibilty of comparing a trace segment from one signature or toolmark to any one in the second signature or toolamrk.

Neither of above sets of correlation are allowed to include the maximum correlation window as identified earlier.

Therefore the assumption is that if two toolmarks or signatures match each other the same-shift correlations would be larger than the different-shift windows,
and if they are not a match the correlations in the two sets will be very similar.

###U Statistic:

This is computed from the joint rank of all correlations of both the same and different shift samples. As given by \citet{hadler}

Null Hypothesis: If the toolmarks were not match i.e not made by the same tool.

Let ns and nd be the number of same shift and different shift windows
$$N = n_{s} + n_{d}$$

The mann whitney U statistic is given by
$$U =\sum^{ns}_{i=1}R_{s}\left( i\right)$$

with the standardized version which includes provision for rank ties

$$\overline{U}= \dfrac{U-M}{\sqrt{V}}$$

where prior to normalization the U-statistic has the mean as

$$M = n_{s}\left(\dfrac{N+1}{2}\right)$$

and variance

$$V = \dfrac{n_{s}n_{d}}{N\left(N-1\right)}\left[\Sigma^{n_{s}}R_{s}\left(i\right)^{2}+\Sigma^{n_{d}}R_{d}\left(j\right)^{2}\right] -\dfrac{n_{s} n_{d}\left(N+1\right)^{2}}{4\left(N-1\right)}$$

## Potential limitations of the Chumbley Score

\hh{in the application to striation marks on bullet lands: smaller in width and curved - need to adjust parameter settings; something similar is done in \citet{afte-chumbley} for toolmark comparisons of slip-joint pliers. XXX work in }

Bullets are much smaller in length, width, are not flat and curved in the cross-sectional topography as opposed to tools like screw driver tips which produced longer and pronounced markings. This means the makings made by barrels in comparison to toolmarks may have a problem in distinctiveness. The majority of Bullet profiles and signatures extracted by procedures mentioned by  \citet{aoas} are almost 1/4 th the size of toolmarks as used by \citet{chumbley} or even smaller. Striations on Bullets are made on their curved surfaces, whereas the algorithm developed by \citet{chumbley} and \citet{hadler} has only been tested for flatter and wider surfaces which have negligible curvature. Therefore, using methods proposed for toolmarks may need adaptation in order to give tangible results for bullets. Moreover, in order to to get flat bullet signatures and remove the curvatures some kind of smoothing needs to be applied as a pre-step which needs further investigation as to whether the level of smoothing does effect the working of the algorithm on Bullets.

Also when in the optimization step, the Window of optimization for bullets will be shorter as the signatures are smaller. The idea is to keep the number of windows of optimization sufficiently large, which means shorter Trace segments (partition of signature or toolmark with length = size of window of optimization) that lets us compare smaller segments of one signature to another.

This introduces a problem as, if we go too small in the window of optimization, the unique features of the trace segments are lost and seem similar, while too large sizes vastly reduces the weight of small features that would otherwise uniquely classify a signature and hence identify the region of agreement.

Thus the Window of Optimization has a direct influence on whether we are
Falsely rejecting the null when it is true (Type I) or Falsely accepting the null when it is false (Type II) making identification of the optimum size of window of optimization very important.

## Testing setup

Following on similar lines to the setup of toolmarks, the first step here is to first identify what difference does different window sizes of optimization and the validation step have, when adapting the toolmark method to bullets.

The marking made on bullets are smaller than toolmarks and is also less wider. The idea is to find out possible areas of error while adapting the score based method proposed for toolmarks, using cross-validation setup to identify appropriate parameter settings for (a) signatures and (b) profiles directly

### Signatures

Signatures of lands for all Hamby-44 and Hamby-252 scans made available through the NIST ballistics database \citep{nist} were considered. Both of these sets of scans are part of the larger Hamby study \citep{hamby} and each consist of twenty known bullets (two each from ten consecutively rifled Ruger P85 barrels) and fifteen questioned bullets (each matching one of the ten barrels). Ground truth for both of these Hamby sets is known and was used to assess correctness of the tests results.

Bullet signatures being compared at this time are therefore from the Hamby 44 and Hamby 252 data. The database setup and pre-processing system used for choosing the Bullet signatures are as described by \citet{aoas}. In order to choose the bullet signatures we first filter out Land_id for Profiles from the Hamby 44 and Hamby 252 data and remove all NA values. Then run_id = 3 is chosen as the signatures generated from this run_id give the closest match. Different run_id's have some different settings for generating the signatures.(The level of smoothing does not seem to be one of them)

The bullet signatures when generated by this process already includes a loess smoothing. Therefore, the coarseness factor is set to 1 while running the chumbley non random algorithm for comparing different optimization windows.The algorithm generates the same_shift, different_shift, U-Stat and P_value parameters which are then used to calculate the errors associated with different sets of window sizes.

We used the adjusted Chumbley method as proposed in \citet{hadler} and implemented in the R package `toolmaRk` \citep{toolmark} on all pairwise land-to-land comparisons of the Hamby scans (a total of 85,491 comparisons) with the pairwise sets for the comparisons given in the following table.

```{r param, echo=FALSE, results='asis'}
param_setting<- all %>% select(wo, wv) %>% group_by(wo, wv) %>% tally() %>% select(wo,wv)
#param_setting<- data.frame(matrix(param_setting, nrow = 16))
#t(param_setting)
#param_setting %>% tally() 
library(kableExtra)
knitr::kable(t(param_setting)[,1:17], digits=0, caption = "Overview of parameter settings used for optimization and validation windows for bullet land signatures.", format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down", "hold"))

knitr::kable(t(param_setting)[,18:33], digits=0, format = "latex" ,booktabs=TRUE) %>% kableExtra::kable_styling(latex_options = c("scale_down","hold"))

# library(xtable)
# strCaption <- "Overview of parameter settings used for optimization and validation windows for bullet land signatures."
#   
#   # set up xtable output
# print(xtable(t(param_setting)[,1:17], 
#              caption = strCaption, label = ""),
#       include.rownames = TRUE,
#       include.colnames = FALSE,
#       caption.placement = "top",
#       comment=FALSE,
#       floating=FALSE
#       )
# print(xtable(t(param_setting)[,17:33],label = ""),
#       include.rownames = FALSE,
#       include.colnames = FALSE,
#       caption.placement = "bottom",
#       comment=FALSE,
#       floating=TRUE
#       #table.placement = "!ht"
#       )
```

### Profiles

The profiles are cross-sectional values of the the bullet striation mark which are chosen at an optimum height (x as used by \citet{aoas}). This x or height is not a randomly chosen level. The rationale behind the choice has been explained by \citet{aoas}. A region is first chosen where the cross-correlation seems to change very less and in this region an optimum height is chosen. The profiles generally resemble a curve which is more or less similar to a quadratic curve (a quadratic fit to the raw data values of the profile is not an exact fit but it does show a similar trend). Profiles are the set of raw values representing the striation marks, and signatures are generated from these by removal of the inherent curvature and applying some smoothing (the signatures generated by \citet{aoas} use a loess function for smoothing).

Similar to signatures the run_id = 3 was used when applying the chumbley algorithm using the database setup given by \citet{aoas} of Hamby-44 and Hamby-252 datasets, on the profiles. The run_id not only defines the level of smoothing but also signifies the chosen height at which the profiles were selected initially. Another important aspect is the range of horizontal values (which is referred to as the y values in \citet{aoas}) in the signatures. These have already been pre-processed in the database to not include any grooves.

Therefore for the sake of comparison the run_id = 3 is still chosen so as to ensure that the horizontal values remain the same as that of the signatures. This also gives us profiles with the grooves removed.

The idea therefore is to first use these raw values of the profile directly in the chumbley algorithm, and see how the algorithm performs for different coarseness values (smoothing parameter as referred in the function lowess used in the chumbley algorithm)

```{r }

#knitr::kable(a27, digits=0, caption = "Window of Optimization 320 and Window of Validation 50", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"))

```

```{r}

#knitr::kable(a27, digits=0, caption = "Window of Optimization 120 and Window of Validation 50", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"))
```

```{r confusion}
library(kableExtra)
a27<- all %>% filter(wo==200, wv==30) %>% xtabs(data= ., ~signif+match)
a27<- as.data.frame(a27)
a27$Type<- c("True Negative","False Positive (Type I)","False Negative (Type II)","True Positive")

a28<- all %>% filter(wo==120, wv==30) %>% xtabs(data= ., ~signif+match)
a28<- as.data.frame(a28)
a28$Type<- c("True Negative","False Positive (Type I)","False Negative (Type II)","True Positive")

a29<- all %>% filter(wo==80, wv==30) %>% xtabs(data= ., ~signif+match)
a29<- as.data.frame(a29)
a29$Type<- c("True Negative","False Positive (Type I)","False Negative (Type II)","True Positive")

knitr::kable(list(a27,a28,a29), digits=0, caption = "Confusion Table for different optimization window sizes with validation window size as 30", format = "latex" ,booktabs=TRUE)%>% kableExtra::kable_styling(latex_options = c("hold"), "striped", full_width = F) %>% group_rows("Size of Optimization Window = 200",1,5) %>% group_rows("Size of Optimization Window = 120", 6, 10) %>%  group_rows("Size of Optimization Window = 80", 11,14)
```

# Results

### Signatures

Figure \ref{fig:type2} gives an overview of type II error rates observed when varying the window size in the optimization step.

>>>>

---
title: "Adaption of the Chumbley Score to matching of bullet striation marks"
authors:
- affiliation: Department of Statistics, Iowa State University 
  name: Ganesh Krishnan
  thanks: The authors gratefully acknowledge ...
- affiliation: Department of Statistics and CSAFE, Iowa State University 
  name: Heike Hofmann
biblio-style: apsr
output:
  pdf_document:
    citation_package: natbib
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
    template: template.tex
  html_document: default
blinded: 0
keywords:
- Bullets, Signatures, Lands, Errors, Known match, Known non-match, Comparisons
bibliography: bibliography
<!-- abstract: null -->
---

@@

<!-- \tableofcontents -->

```{r setup, echo = FALSE, message = FALSE, warning = FALSE}
library(knitr)
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  collapse = TRUE,
  comment = "",
  fig.height = 6,
  fig.width = 6,
  fig.align = "center",
  out.width= '\\textwidth',
  cache = FALSE,
  fig.path='figures/',
  echo=FALSE,
  cache=TRUE
)
options(knitr.table.format = "latex")
library(tidyverse)
```
# Introduction and Results
